gazeHMM: Parameter recovery simulation

### 2020-05-06 simulation/preregistration_parameter_recovery.Rmd
add parameter recovery script

---
title: "gazeHMM: Parameter recovery simulation"
author: "Malte Lüken, Simon Kucharsky, & Ingmar Visser"
date: "5 5 2020"
output: pdf_document
---

In this document, we are preregistering a simulation study to estimate the parameter recovery of a hidden Markov model (HMM) which is part of our recently developed algorithm gazeHMM. The purpose of this algorithm is to classify eye movements into distinct events (e.g., fixations). The full algorithm will be available on Github upon publication.

In general, gazeHMM consists of three steps: First, the raw gaze data is preprocessed. Missing samples are identified, the data are filtered, and for each sample the velocity, acceleration, and difference in angle to the next sample is calculated. Second, the HMM estimates the parameters of response, transition, and initial state models for the specified eye movement events. Moreover, the HMM computes the posterior probability of each sample for belonging to the events. Each sample is labeled as the event with the highest posterior probability. Third, a postprocessing routine relabels samples that are violating theoretical and practical constraints. Since the HMM might not model the data perfectly, this step improves classification performance. This simulation study will only consider the parameter recovery of the HMM but not the performance of the preprocessing and postprocessing routines of the algorithm.

# Model description

The HMM estimates unobservable states that have generated the data. In the context of eye-tracking, each state corresponds to an eye movement event. In gazeHMM, the model can have between two and four states. State one always corresponds to fixations, state two always to saccades, state three to postsaccadic oscillations (PSOs), and state four to smooth pursuits. Moreover, the HMM is multivariate but conditionally independent on the states. Thus, it has three independent response models for every state describing velocity, acceleration, and angle. The velocity and acceleration response models for all four states use two-parameter gamma distributions (shape and scale parametrization). The angle response model for the fixation state uses a uniform distribution, while the other three states use a two-parameter von-Mises distribution (mean and concentration parametrization) to model angle. Both the transition model and the initial state model use a multinomial distribution (with the logit and the identity link function, respectively). Note that no covariates are included in the model, only intercepts for the parameters are estimated.

Depending on the number of states, the HMM has a different number of parameters. The response models of the HMM contain six parameters per state. The transition models have $k^k$ parameters in total with k being the number of states. The initial state models have one paramter per state. In sum, HMMs with two, three, and four states have X, X, and X paramters. The denotation of each parameter is shown in Table 1.

*Optimization*

@@

|Parameter|Description|
|:--------|:----------|
|$\pi$   |Initial probabilities for state *i*|
|$a_{ij}$ |Transition probability from state *i* to *j*|
|$

# Model parameter recovery

The HMM will always start with a uniform distribution to estimate the initial state and state transition probabilites. To generate random starting values for the estimation of shape, scale, and concentration parameters, we will use gamma distributions with the true parameter as the shape parameter and a scale parameter of $\beta = 1$. Mean parameters of the von-Mises distribution will always start at their true value.

In the first part of the simulation study, we will estimate parameter recovery under ideal condititions. These imply an average sample size of $N = 2500$ samples per data set, no noise added to the data, and no missing samples. The parameters that we will use to generate the data are displayed in Table 2. We will vary one parameter at a time on the given interval and set the other parameters to their fixed value. This simulation will result in $D_{total} = 100 \times 3 \times 20 = 6000$ recoveries.

In the second part, we will vary the sample size of the generated data and the amount of noise added to it. For sample sizes of $N \in \{500, 2500, 10000\}$, we will generate data sets and add white noise. To velocity and acceleration, we will add gaussian noise with $\mu = 0$ and $\sigma \in [1,25]$. Noise from a von-Mises distribution with $\mu = 0$ and $\kappa \in 1/[0.1,10]$ will be added to angle. This simulation will result in $D_{total} = 100 \times 3 \times 3 = 900$ recoveries.

In the third part, we will increase the variation in the starting values used for parameter estimation. For the shape, scale, and concentration parameters, we will increase the scale parameters of the gamma distributions to $\beta \in \{2, 4, 8\}$ all at the same time. The sample size will be $N = 2500$ and no noise will be added to the data. This simulation will result in $D_{total} = 100 \times 3 \times 3 = 900$ recoveries.

In the last part, we will set intervals of the generated data to be missing. The size of the missing data interval will be $m \in [1,100]$ and the number of intervals will be $n_{miss} \in \{1, 3, 5\}$. The sample size will be $N = 2500$ and no noise will be added to the data. This simulation will result in $D_{total} = 100 \times 3 \times 3 = 900$ recoveries.

Table 2

|State|Parameter|Interval|Fixed|
|-----|---------|--------|-----|
|1-4|$p$*|-|$1/k$|
|1-4|$a_{i=j}$|[.01,.99]|0.9|
|1-4|$a_{i \neq j}$|$(1-a_{i=j})/(k-1)$|$0.1/(k-1)$|
|1|$\alpha_{vel}$|[1,5]|3|
|1|$\beta_{vel}$|[0.5,1]|0.75|
|1|$\alpha_{acc}$|[1,5]|3|
|1|$\beta_{acc}$|[0.05,0.5]|0.15|
|1|$a$*|-|-|
|1|$b$*|-|-|
|2|$\alpha_{vel}$|[1,5]|3|
|2|$\beta_{vel}$|[5,15]|10|
|2|$\alpha_{acc}$|[1,5]|3|
|2|$\beta_{acc}$|[1,5]|3|
|2|$\mu$*|-|0|
|2|$\kappa$|$1/[0.1,10]$|1|
|3|$\alpha_{vel}$|[1,5]|3|
|3|$\beta_{vel}$|[1,5]|3|
|3|$\alpha_{acc}$|[1,3]|2|
|3|$\beta_{acc}$|[1,3]|2|
|3|$\mu$*|-|$\pi$|
|3|$\kappa$|$1/[0.1,10]$|1|
|4|$\alpha_{vel}$|[1,5]|3|
|4|$\beta_{vel}$|[1,2]|1.5|
|4|$\alpha_{acc}$|[1,5]|3|
|4|$\beta_{acc}$|[0.05,0.25]|0.15|
|4|$\mu$*|-|0|
|4|$\kappa$|$1/[0.1,10]$|1|

*Note*. Parameters marked with * will not be varied but always set to their fixed values. Parameters *a* and *b* (the minimum and maximum of the uniform distribution) are always obtained from the data.

>>>>

---
title: "gazeHMM: Parameter recovery simulation"
author: "Malte Lüken, Simon Kucharský, & Ingmar Visser"
affiliation: "University of Amsterdam"
date: "05.05.2020"
output: pdf_document
---

In this document, we are preregistering a simulation study to estimate the parameter recovery of a hidden Markov model (HMM) which is part of our recently developed algorithm gazeHMM. The purpose of this algorithm is to classify eye movements into distinct events (e.g., fixations and saccades). The full algorithm will be available on Git hub upon publication.

In general, gazeHMM consists of three steps: First, the raw gaze data is preprocessed. Missing samples are identified, the data are filtered, and for each sample the velocity, acceleration, and difference in angle to the next sample is calculated. Second, the HMM estimates the parameters of response, transition, and initial state models for the specified eye movement events. Moreover, the HMM computes the posterior probability of each sample for belonging to the events. Each sample is labeled as the event with the highest posterior probability. Third, a postprocessing routine relabels samples that are violating theoretical and practical constraints. This simulation study will only consider the parameter recovery of the HMM but not the performance of the preprocessing and postprocessing routines of the algorithm.

# Model Description

The HMM estimates non-observable states that have generated the data. In the context of eye-tracking, each state corresponds to an eye movement event. In gazeHMM, the model can have between two and four states. State one always corresponds to fixations, state two always to saccades, state three to postsaccadic oscillations (PSOs), and state four to smooth pursuits. Moreover, the HMM is multivariate but conditionally independent on the states. Thus, it has three independent response models for every state describing velocity, acceleration, and angle. The velocity and acceleration response models for all four states use two-parameter gamma distributions (shape and scale parametrization). The angle response model for the fixation state uses a uniform distribution, while the other three states use a two-parameter von-Mises distribution (mean and concentration parametrization) to model angle. Both the transition model and the initial state model use a multinomial distribution (with the logit and the identity link function, respectively). Note that no covariates are included in the model, only intercepts for the parameters are estimated.

The HMM uses an expectation-maximization algorithm to estimate parameters under the hidden states. The parameters of each response model (except for the uniform distribution) are estimated with a spectral projected gradient method and Barzilai-Borwein step lengths.

# Model Parameter Recovery

The HMM will always start with a uniform distribution to estimate the initial state and state transition probabilities. To generate random starting values for the estimation of shape, scale, and concentration parameters, we will use gamma distributions with the true parameter as the shape parameter and a scale parameter of $\beta = 1$. Mean parameters of the von-Mises distribution will always start at their true value.

In the first part of the simulation study, we will estimate parameter recovery under ideal conditions. These imply an average sample size of $N = 2500$ samples per data set, no noise added to the data, and no missing samples. The parameters that we will use to generate the data are displayed in Table 1. We will vary one parameter at a time on the given interval and set the other parameters to their fixed value. For two states, 10 parameters will be varied one at a time, for three states 15, and for four states 20 parameters. This simulation will result in $D_{total} = 100 \times (10+15+20) = 4500$ recoveries.

In the second part, we will vary the sample size of the generated data and the amount of noise added to it. The model parameters will be set to the fixed values from Table 1. For sample sizes of $N \in \{500, 2500, 10000\}$, we will generate data sets and add white noise. To velocity and acceleration, we will add Gaussian noise with $\mu = 0$ and $\sigma \in [1,25]$. Noise from a von-Mises distribution with $\mu = 0$ and $\kappa \in 1/[0.1,10]$ will be added to angle. This simulation will result in $D_{total} = 100 \times 3 \times 3 = 900$ recoveries.

In the third part, we will increase the variation in the starting values used for parameter estimation. The model parameters will be set to the fixed values from Table 1. For the shape, scale, and concentration parameters, we will increase the scale parameters of the gamma distributions to $\beta \in \{2, 4, 8\}$ all at the same time. The sample size will be $N = 2500$ and no noise will be added to the data. This simulation will result in $D_{total} = 100 \times 3 \times 3 = 900$ recoveries.

In the last part, we will set intervals of the generated data to be missing. The model parameters will be set to the fixed values from Table 1. The size of the missing data interval will be $m \in [1,100]$ and the number of intervals will be $n_{miss} \in \{1, 3, 5\}$. The sample size will be $N = 2500$ and no noise will be added to the data. This simulation will result in $D_{total} = 100 \times 3 \times 3 = 900$ recoveries.

@@

|State|Parameter|Interval|Fixed|Description|
|-----|---------|--------|-----|-----------|
|1-4|$p_i^+$|-|$1/k$|Initial state probability for starting in state *i*|
|1-4|$a_{i=j}$|[.01,.99]|0.9|Transition probability for staying in the previous state *i*|
|1-4|$a_{i \neq j}$|$(1-a_{i=j})/(k-1)$|$0.1/(k-1)$|Transition probability for switching to from state *i* to a different state *j*|
|1|$\alpha_{vel}$|[1,5]|3|Shape parameter of the velocity gamma distribution|
|1|$\beta_{vel}$|[0.1,0.6]|0.35|Scale parameter of the velocity gamma distribution|
|1|$\alpha_{acc}$|[1,5]|3|Shape parameter of the acceleration gamma distribution|
|1|$\beta_{acc}$|[0.05,0.25]|0.15|Scale parameter of the acceleration gamma distribution|
|1|$a^+$|-|0|Minimum of the uniform distribution|
|1|$b^+$|-|$2\pi$|Maximum of the uniform distribution|
|2|$\alpha_{vel}$|[1,5]|3||
|2|$\beta_{vel}$|[5,25]|15||
|2|$\alpha_{acc}$|[1,5]|3||
|2|$\beta_{acc}$|[1,5]|3||
|2|$\mu^+$|-|0|Mean parameter of the von-Mises distribution|
|2|$\kappa$|$1/[0.1,10]$|1|Concentration parameter of the von Mises distribution
|3|$\alpha_{vel}$|[1,5]|3||
|3|$\beta_{vel}$|[1,5]|3||
|3|$\alpha_{acc}$|[1,3]|2||
|3|$\beta_{acc}$|[1,3]|2||
|3|$\mu^+$|-|$\pi$||
|3|$\kappa$|$1/[0.1,10]$|1||
|4|$\alpha_{vel}$|[1,5]|3||
|4|$\beta_{vel}$|[1,2]|1.5||
|4|$\alpha_{acc}$|[1,5]|3||
|4|$\beta_{acc}$|[0.05,0.25]|0.15||
|4|$\mu^+$|-|0||
|4|$\kappa$|$1/[0.1,10]$|1||

*Note*. Parameters marked with $^+$ will not be varied but always set to their fixed value. *k* is the number of states in the model.
